Author disambiguation of large-scale semi-labeled collections of scientific articles

### 2015-08-18 paper.tex
blocking

The simplest and most common strategy for blocking \glnote{add references}
groups signatures together if they share the same surname(s) and the same first
given name initial (e.g., "Doe, J").

\vspace{3cm}

If the goal of the algorithm is to disambiguate millions of signatures and the algorithm
complexity is $\Omega{(n^2)}$ in terms of the number of signatures due to pairwise
comparisons, the disambiguation might take weeks or months to finish. In order to prevent
that, we \textit{block} the signatures \citep{fellegi69}. The simplest strategy groups the
signatures together if they share the same surnames and first given name initial (e.g.,
"Wang, Z"). As the clustering will be performed only within the
blocks, such strategy might prevent the further steps of the algorithm from assigning
correctly two signatures to the same author. We extracted from our training set all pairs
where the signatures belong to different blocks and the same author after the last name and
first initial strategy was used. We observed that we can classify majority of these pairs
into few categories:
\begin{enumerate}
  \item Author has multiple surnames and on some of his signatures all the surnames but the
  last one are contained within given names (e.g., "Martinez Torres, A." and "Torres, A.
  Martinez").
  \item Author has multiple surnames and on some of his signatures only the first surname is
  present (e.g., "Smith-Jones, A." and "Smith, A.")
  \item Author comes from a country that doesn't use latin alphabet and there are different
  ways of transcripting his name or his name sometimes contains a typo on the signatures
  (e.g., "Mueller, R." and "Muller, R.", "Tchaikovsky, P." and "Czajkowski, P.").
  \item Author sometimes uses on the signature a different name than his first one (e.g.,
  "Smith, Jack" and "Smith, A.J.")
  \item Author changed his surname (e.g., because of marrying somebody).
\end{enumerate}

We tackled first four categories\msnote{Should I mention why not all of them?} and
implemented a blocking strategy that produces less pairs with signatures from the same
author but different blocks, and the blocks output by the algorithm are small. Before
performing blocking, the names on the signatures are \textit{normalized}: the accents are
removed (e.g., "Jabłoński, Ł" $\rightarrow$ "Jablonski, L") and common affixes are removed
(e.g., "van der Waals, J. D."  $\rightarrow$ "Waals, J. D.")\msnote{Better explanation of
"affix" needed?}.
\begin{algorithm}
  \caption{Blocking using phonetic algorithm}
  \begin{algorithmic}[1]
    \newcommand{\myindent}[1]{
        \newline\makebox[#1cm]{}
    }
    \Require{$S$ - set of signatures. The names on the signatures should be
             \textit{normalized},
             \myindent{0}$phonetic\_surnames$ - functions that outputs phonetic
             representation of surnames from \myindent{1.38}given signature,
             \myindent{0}$given\_names$ - function that ouputs all the given names from
             given signature.}

    \Function{Phonetic blocking}{$S$}
      \State $blocks\gets \emptyset$
      \Comment{Blocks are indexed by a phonetic representation of a surname}
      \ForAll{$s$ in $S$ with single surname}
         \myindent{0.92}assign $s$ to the block of $phonetic\_surnames(s)[0]$
      \EndFor
      \ForAll{$s$ in $S$ with multiple surnames}
         \If{block of $phonetic\_surnames(s)[-1]$ contains a signature $t$
         \myindent{1.84}where $phonetic\_surnames(s)[0] = given\_names(t)[-1]$}
         \myindent{1.38}assign $s$ to the block of $phonetc\_surnames(s)[-1]$
         \ElsIf{block of $phonetic\_surnames(s)[0]$ exists}
          \myindent{1.38}assign $s$ to the block of $phonetic\_surnames(s)[0]$
         \Else
          \myindent{1.38}assign $s$ to the block of $phonetic\_surnames(s)[-1]$
         \EndIf
      \EndFor
      \Return $blocks$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\msnote{Should I explain that if the block does not exist, an empty one is created?}

After this algorithm is finished, all the blocks from the result of the function above can
be additionally split over the first given name initial.

\subsection{Learning a similarity function}
\glnote{By Gilles}

@@

\subsection{Semi-supervised clustering}
\label{methods:clustering}

@@

>>>>

The simplest and most common strategy for blocking \glnote{add references}
groups signatures together if they share the same surname(s) and the same first
given name initial (e.g., "Doe, J"). Despite satisfying performance, such
a simple strategy fails in several cases, as listed below for
pairs of signatures that should be clustered together but that are put into distinct
blocks:

\begin{enumerate}
  \item There are different
  ways of transcripting an author name or some of signatures contain a typo
  (e.g., "Mueller, R." and "Muller, R.", "Tchaikovsky, P." and "Czajkowski, P.").

  \item An author has multiple surnames and, on some of his signatures, all the surnames but the
  last one are contained within given names (e.g., "Martinez Torres, A." and "Torres, A.
  Martinez").

  \item An author has multiple surnames and, on some of his signatures, only the first surname is
  present (e.g., "Smith-Jones, A." and "Smith, A.")

  \item An author as multiple given names and they are not always mentioned on his signatures (e.g.,
  "Smith, Jack" and "Smith, A. J.")

  \item An author changed his surname (e.g., because of marrying somebody).
\end{enumerate}

To account for these issues we propose instead to block signatures based on the
phonetic representation of the normalized surname. Normalization consists in
stripping accents (e.g., "Jabłoński, Ł" $\rightarrow$ "Jablonski, L") and name
affixes that inconsistenly appear in signatures (e.g., "van der Waals, J. D."
$\rightarrow$ "Waals, J. D."), while phonetization is based either on the
Double Metaphone \citep{doublemetaphone}, the NYSIIS \citep{nysiis} or the
Soundex \citep{Soundex} algorithms. Together, these processing steps allow to
group most name variants of a same person in the same block while not
significantly increasing computational complexity, thereby solving case 1.

In the case of multiple surnames (cases 2 and 3), we further propose to block
signatures in two phases. In the first phase, all the signatures with  one
surname are clustered together. Every different surname token creates a new
block. In the second phase, the signatures  with multiple surnames are compared
with the blocks for the first and last surname.  If the first surnames of an
author were already used as the last given names on some of the signatures, the
new signature is assigned to the block of the last surname (case 2). Otherwise,
the signature is assigned to the block of the first surname (case 3). Finally,
to prevent the creation of too large blocks, signatures are further divided
along their first given name initial.  Cases 4 and 5 are therefore not
explicitly handled.

\glnote{I am not sure we need an explicit algorithm, but adding a few examples would help.}

\subsection{Learning a similarity function}
\label{methods:similarity}
\glnote{By Gilles}

@@

\subsection{Semi-supervised clustering}
\label{methods:clustering}
\glnote{By Gilles}

@@

\glnote{To proofread / improve}
